### 2021-11-17 documents/floats/yukreg.tex
better overview table
md
'to go down'

\begin{table}
\centering
\caption[Regular \yukpa verbs]{Regular \yukpa verbs \parencites[72, 76]{largo2011yukpa}[139]{meira2006syntactic}}
\label{tab:yukreg}
\begin{tabular}[t]{@{}llll@{}}
\mytoprule
{} &  \qu{to fall} & \qu{to sleep} & \qu{to wash self} \\
\midrule
\gl{1} &  \obj{j-ata-} &  \obj{jɨ-nɨ-} &     \obj{j-otum-} \\
\gl{2} &  \obj{m-ata-} &  \obj{mɨ-nɨ-} &     \obj{m-otum-} \\
\gl{3} &  \obj{n-ata-} &  \obj{nɨ-nɨ-} &     \obj{n-otum-} \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\begin{table}
\centering
\caption[Regular \yukpa verbs]{Regular \yukpa verbs \parencites[139]{meira2006syntactic}[72, 76]{largo2011yukpa}}
\label{tab:yukreg}
\begin{tabular}[t]{@{}llll@{}}
\mytoprule
{} &  \qu{to fall} & \qu{to wash self} & \qu{to sleep} \\
\midrule
\gl{1} &  \obj{j-ata-} &     \obj{j-otum-} &  \obj{jɨ-nɨ-} \\
\gl{2} &  \obj{m-ata-} &     \obj{m-otum-} &  \obj{mɨ-nɨ-} \\
\gl{3} &  \obj{n-ata-} &     \obj{n-otum-} &  \obj{nɨ-nɨ-} \\
\bottomrule
\end{tabular}
\end{table}

### 2021-11-17 documents/sections/verbs.tex
'to go down'

\subsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{k-} in \PPek \pcref{sec:pekodian} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
The resistance against the former extension was subsequently overcome in \bakairi; its fate in \ikpeng is unknown.
When \akuriyo extended \obj{k-}, the verb already had an irregularly inflected first person form with \obj{p-}, inherited from \PTir.

At first sight, it may seem that it also was affected by the independent extensions of \obj{j-} in \carijo \exref{gowhere.car-32} and \yukpa \exref{gowhere.yuk-14}.

\pex<gowhere>
\a<car-32> \carijo \perscommpar{David Felipe Guerrero}\\
\begingl
\gla irə wat͡ʃinakano tae j-ehɨtə-e//
\glb \gl{inan}.\gl{ana} body.of.water along.bounded \gl{1}-go.down-\gl{npst}//
\glft \qu{…I go down through that guachinacán.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}\gl{pro} \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
However, a look at the comparative picture suggests a much more complicated situation.
\cref{tab:godown} shows all attested forms, including verb class membership where applicable; parenthesized forms indicate uncertainty about cognacy status.
It turns out that while a form \rc{ɨpɨtə} can be reconstructed to \PC, different (proto-)languages do not agree about the class of this verb.
Its reflexes in languages that preserve the split-\gl{s} system are split fairly evenly between \gl{s_a_} and \gl{s_p_}.

@@

\pex<waygodown>\wayana \parencite[][200]{wayanatavares2005}
\a<way-73>
\begingl
\glpreamble ïwïptëë//
\gla ɨ-w-ɨptə-rɨ//
\glb \gl{1}-\gl{s_a_}-go.down-\gl{nmlz}//
\glft \qu{my going down}//
\endgl
\a<way-71>
\begingl
\gla əw-ɨptə-k//
\glb \gl{2}-go.down-\gl{imp}//
\glft \qu{Go down!}//
\endgl
\xe
%
Its causativized form is \obj{ɨptə-ka} \parencite[255]{wayanatavares2005}; the restriction of \rc{-ka} to \gl{s_p_} verbs in \PC \parencite{gildea2019overview} points to \gl{s_p_} membership.
These patterns lead us to posit the hypothesis that the verb was a regular member of the \gl{s_p_} class in pre-\wayana, but partially switched to the \gl{s_a_} class.
This in turn implies that all reflexes of this verb with \gl{s_a_} status at some point switched from the \gl{s_p_} class.

\wayana-external comparative evidence supports this hypothesis:
The \arara causativized form is \obj{eniptoŋ} \parencite[66]{alves2017arara}, and \kalina has a cognate form \obj{enɨʔto} \parencite[263]{courtz2008carib}; \obj{onɨʔto} \qu{to go down} in \cref{tab:godown} is a detransitivized form thereof, lit.\ \qu{to get oneself down}.
Both causativized forms contain a reflex of the transitivizer \rc{en-}, which was usually found with \gl{s_p_} verbs \parencite{gildea2019overview}.
Besides the irregular first person \obj{p-}, \trio \obj{ɨhtə} shows other irregularities, in particular in its causativized forms \parencite[263]{triomeira1999}.
Thus, it seems that this verb was originally \gl{s_p_}, but then switched its class in four and a half languages of the family, for so far unknown reasons.

These circumstances make it impossible to answer the question of whether \qu{to go down} was affected by the extensions in \PWai, \PTir, \carijo, and \yukpa.
For \PTir, we cannot establish a relative chronology of the verb class change, the introduction of the idiosyncratic marker \rc{p-}, and the extension of \rc{t-}.
For \PWai, we lack knowledge not only about the first person form, but even about class membership (in \hixka).
For \carijo and \yukpa, we cannot know whether the verb potentially switched class before the breakdown of the entire split-\gl{s} system.
While there is no language-internal evidence for such a switch, \qu{to go down} does have an inclination for class switches; in the case of \carijo, that could have already happened at the \PTar stage.
In all four cases, it is possible that the verb had \gl{s_a_} status at the time of the extension, resisting and keeping its old marker, but it is also possible that it was not even a potential target due to its \gl{s_p_} status.
On the other hand, the class change must have happened before the split-up of \trio and \akuriyo, and therefore this verb resisted the extension of \akuriyo \obj{k-}.
Likewise, it seems very likely that the class change took place before the extension of \PPek \rc{k-}.
Otherwise, the newly-turned-\gl{s_a_} verb would have taken on archaic and lexically heavily restricted \rc{w-}, either in \PTar, \PXin, or \arara.

>>>>

\subsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{k-} in \PPek \pcref{sec:pekodian} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
The resistance against the former extension was subsequently overcome in \bakairi; its fate in \ikpeng is unknown.
When \akuriyo extended \obj{k-}, the verb already had an irregularly inflected first person form with \obj{p-}, inherited from \PTir.
At first sight, it may seem that it also was affected by the two independent extensions of \obj{j-} in \carijo \exref{gowhere.car-32} and \yukpa \exref{gowhere.yuk-14}.

\pex<gowhere>
\a<car-32> \carijo \perscommpar{David Felipe Guerrero}\\
\begingl
\gla irə wat͡ʃinakano tae j-ehɨtə-e//
\glb \gl{inan}.\gl{ana} body.of.water along.bounded \gl{1}-go.down-\gl{npst}//
\glft \qu{…I go down through that guachinacán.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}\gl{pro} \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
However, a broader comparative perspective reveals a much more complicated story.
\cref{tab:godown} shows all attested cognates of \rc{ɨpɨtə}, including verb class membership where applicable; parenthesized forms indicate uncertainty about cognacy status.
It turns out that while a form \rc{ɨpɨtə} can be reconstructed to \PC, different (proto-)languages do not agree about the class of this verb.
Its reflexes in languages that preserve the split-\gl{s} system are distributed fairly evenly between \gl{s_a_} and \gl{s_p_}.

@@

\pex<waygodown>\wayana \parencite[][200]{wayanatavares2005}
\a<way-73>
\begingl
\glpreamble ïwïptëë//
\gla ɨ-w-ɨptə-rɨ//
\glb \gl{1}-\gl{s_a_}-go.down-\gl{nmlz}//
\glft \qu{my going down}//
\endgl
\a<way-71>
\begingl
\gla əw-ɨptə-k//
\glb \gl{2}-go.down-\gl{imp}//
\glft \qu{Go down!}//
\endgl
\xe
%
Its causativized form is \obj{ɨptə-ka} \parencite[255]{wayanatavares2005}; the restriction of \rc{-ka} to \gl{s_p_} verbs in \PC \parencite{gildea2019overview} points to \gl{s_p_} membership.
These patterns lead us to posit the hypothesis that the verb was a regular member of the \gl{s_p_} class in pre-\wayana, but partially switched to the \gl{s_a_} class, reflected only in its \gl{1+2} prefix and the \gl{s_a_} class marker.
This in turn implies that reflexes of this verb with \gl{s_a_} membership in other languages fully switched from \gl{s_p_} at some point.

\wayana-external comparative evidence supports this hypothesis:
The \arara causativized form is \obj{eniptoŋ} \parencite[66]{alves2017arara}, and \kalina has a cognate form \obj{enɨʔto} \parencite[263]{courtz2008carib}; \obj{onɨʔto} \qu{to go down} in \cref{tab:godown} is a detransitivized form thereof, lit.\ \qu{to get oneself down}.
Both causativized forms contain a reflex of the transitivizer \rc{en-}, which was usually found with \gl{s_p_} verbs \parencite{gildea2019overview}.
Besides the irregular first person \obj{p-}, \trio \obj{ɨhtə} shows other irregularities, in particular in its causativized forms that also feature a reflex of \rc{en-} \parencite[263]{triomeira1999}.
Thus, it seems that this verb was originally \gl{s_p_}, but then switched its class in four and a half languages of the family, for so far unknown reasons.

These circumstances make it impossible to answer the question of whether \qu{to go down} was affected by the extensions in \PWai, \PTir, \carijo, and \yukpa.
For \PTir, we cannot establish a relative chronology of the verb class switch, the introduction of the idiosyncratic marker \rc{p-}, and the extension of \rc{t-}.
For \PWai, we lack knowledge not only about the first person form, but even about class membership.
Similarly, for \carijo and \yukpa we cannot know whether the verb potentially switched class before the breakdown of the entire split-\gl{s} system.
While there is no language-internal evidence, it is clear that \qu{to go down} does have an inclination to switch classes; in the case of \carijo, that could have already happened at the \PTar stage.
In all four cases, it is possible that the verb had \gl{s_a_} status at the time of the extension, resisting it and preserving the old prefix, but it is also possible that it was not even a potential target due to its \gl{s_p_} status at the time.
On the other hand, the class switch happened before the split-up of \trio and \akuriyo, and therefore this verb resisted the extension of \akuriyo \obj{k-} as an \gl{s_a_} verb.
Likewise, it seems very likely that the class switch took place before the extension of \PPek \rc{k-}.
Otherwise, the newly-turned-\gl{s_a_} verb would have taken on archaic and lexically heavily restricted \rc{w-}, either in \PTar, \PXin, or \arara.
